---
title: "Mutation Signature Report"
author: "Eric Zhao"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About this Analysis

This pipeline is currently maintained by Eric Zhao, who can be contacted at [ezhao@bcgsc.ca](mailto:ezhao@bcgsc.ca).
It was first developed by [Eric Zhao](www.eyzhao.com) at Canada's Michael Smith Genome Sciences Centre
in January, 2018.

# Analysis of Signatures by WTSI Method

What follows is a *de novo* analysis of mutation signatures using the WTSI framework.
This framework determines signatures using non-negative matrix factorization (NMF),
and is described in detail by 
[Alexandrov et al. (2013)](http://www.cell.com/cell-reports/abstract/S2211-1247(12)00433-0).

Computationally heavy elements of this framework are computed using an upstream pipeline.
1000 NMF iterations are performed for a range of different models, each of which assumes
a different number of latent mutation signatures. We begin by reading in the results from
the analysis.

```{r wtsi_load}
library(tidyverse)
library(cowplot)
library(viridis)

nmf_signatures <- read_tsv('wtsi/combined_signatures.tsv') %>%
  mutate(
    mutation_type = sprintf(
      '%s[%s]%s',
      substr(subtype, 1, 1),
      type,
      substr(subtype, 3, 3)
    )
  )

nmf_exposures <- read_tsv('wtsi/combined_exposures.tsv') %>%
  mutate(sample = gsub('.*?catalog\\/(.*?)\\.tsv', '\\1', sample_names)) %>%
  select(-path, -sample_names)

nmf_metrics <- read_tsv('wtsi/model_selection_metrics.tsv')
```

## Selection of Number of Signatures

The number of mutation signatures is selected based on signature stability and Frobenius reconstruction error.
For this report, we utilize an automated model selection scheme. Signature stability and reconstruction error
are each first renormalized as (x - min(x)) / (max(x) - min(x)) - see code below for implementation details.
Because we are aiming to maximize stability and minimize reconstruction error, we then select the model with
the greatest difference of stability minus reconstruction error.

The mutation signature metrics are plotted for each model below, with the automatically chosen model
highlighted with a black box.

```{r wtsi_plot_metrics, fig.height = 5, fig.width = 5}
metrics_for_plot <- nmf_metrics %>%
  spread(metric, value)

auto_chosen_model = nmf_metrics %>%
  spread(metric, value) %>%
  mutate(
    stability_proportion = (stability - min(stability)) / (max(stability) - min(stability)),
    reconstruction_proportion = (reconstructionError - min(reconstructionError)) / (max(reconstructionError) - min(reconstructionError)),
    combined = stability_proportion - reconstruction_proportion
  ) %>%
  filter(combined == max(combined)) %>%
  filter(row_number() == n()) %>%
  .[['n_signatures']]

metrics_for_plot %>%
  ggplot(aes(
    x = stability,
    y = reconstructionError,
    label = n_signatures
  )) +
  geom_text() +
  geom_label(
    data = metrics_for_plot %>% filter(n_signatures == auto_chosen_model),
    fill = 'black',
    colour = 'white'
  ) +
  labs(
    x = 'Signature Stability',
    y = 'Reconstruction Error'
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = 'bold')
  )

chosen_nmf_signatures <- nmf_signatures %>%
  filter(n_signatures == auto_chosen_model) %>%
  select(signature, mutation_type, proportion)

chosen_nmf_exposures <- nmf_exposures %>%
  filter(n_signatures == auto_chosen_model)
```

## Plot of Mutation Signatures

Based on the above analysis, a `r auto_chosen_model`-signature model has been automatically selected.
We next plot the mutation signatures deciphered using this model. Note that the barplot is coloured
according to the 5' and 3' mutation context, and a legend is provided.

```{r wtsi_plot_signatures, fig.height = 6, fig.width = 10}
colour_df <- data.frame(
  five_prime = c(rep('A', 4), rep('C', 4), rep('G', 4), rep('T', 4)), 
  three_prime = rep(c('A', 'C', 'G', 'T'), 4),
  status = factor(1:16, levels=1:16),
  colour = hsv(h=sort(rep(c(0, 0.3, 0.6, 0.9), 4)), s=rep(c(0.1, 0.2, 0.5, 1), 4), v=rep(c(0.8, 0.8, 0.8, 0.8), 4))
) %>%
  mutate(
    context = paste(five_prime, three_prime, sep='.'),
  )

signature_plots <- chosen_nmf_signatures %>%
  mutate(
    substitution = gsub('.\\[(.*?)\\].', '\\1', mutation_type),
    context = gsub('(.)\\[.*?\\](.)', '\\1.\\2', mutation_type)
  ) %>% left_join(
    colour_df %>% select(context, status), by = 'context'
  ) %>%
  plyr::dlply('signature', function(z) {
  z %>%
    ggplot(aes(
      x = context,
      y = proportion,
      fill = status
    )) + 
    facet_grid(. ~ substitution) +
    geom_bar(stat = 'identity', width = 0.75) +
    theme(
      axis.title.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(size = 10),
      legend.position = 'none',
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    ) +
    scale_fill_manual(values = as.character(colour_df$colour))
})

colour_legend = colour_df %>%
  ggplot(aes(
    x = five_prime,
    y = three_prime,
    fill = status
  )) +
  geom_tile(width = 0.8, height = 0.8) +
  scale_fill_manual(values = as.character(colour_df$colour)) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_text(size = 20),
    legend.position = 'none'
  ) +
  labs(x = "5' context", y = "3' context")

colour_legend_ <- plot_grid(colour_legend, NULL, nrow = 2, rel_heights = c(1, 2.2))

plot_grid(plotlist = signature_plots, ncol = 2, labels = 1:length(signature_plots)) %>%
  plot_grid(colour_legend_, ncol = 2, rel_widths = c(4.5,1))
```

## Comparison against reference signatures

Reference signatures through [COSMIC](http://cancer.sanger.ac.uk/cosmic/signatures) provide a point
of comparison. This allows us to see where our signatures deviate from others discovered in the literature.
The greater the cosine similarity, the more similar a new signature is to a previously discovered one.

```{r wtsi_reference_comparison, fig.height = 4, fig.width = 8}
cosine_similarity <- function(x, y) {
  (x %*% y) / (sqrt(sum(x^2)) * sqrt(sum(y^2)))
}

reference_signatures <- read_tsv('/projects/ezhao_prj/signatures/reference/wtsi_signatures/v2016-04-05/signatures_probabilities_genome.txt') %>%
  gather(signature, reference_probability, -mutation_type) %>%
  rename(reference_signature = signature)

chosen_nmf_signatures %>%
  left_join(reference_signatures, by = 'mutation_type') %>%
  group_by(signature, reference_signature) %>%
  summarise(cosine_similarity = cosine_similarity(proportion, reference_probability)) %>%
  mutate(reference_signature = as.integer(gsub('Signature ', '', reference_signature))) %>%
  ggplot(aes(
    x = reference_signature,
    y = signature,
    fill = cosine_similarity
  )) +
  scale_fill_viridis(trans = scales::exp_trans(base = 10)) +
  geom_tile()
```

## Plot of signature exposures

The mutation signature algorithm simultaneously infers signatures and their relatives contributions
to each cancer genome. The following plot illustrates the these relative contributions, along with
the total mutation burden of each genome.

```{r wtsi_exposures}
nmf_clust <- chosen_nmf_exposures %>%
  select(sample, signature, exposure) %>%
  arrange(sample) %>%
  spread(signature, exposure) %>%
  select(-sample) %>%
  dist %>% hclust

sample_levels_raw <- chosen_nmf_exposures %>% arrange(sample) %>% .$sample %>% unique
sample_levels <- sample_levels_raw[nmf_clust$order]

nmf_exposure_heatmap <- chosen_nmf_exposures %>%
  group_by(sample) %>%
  mutate(exposure = exposure / sum(exposure)) %>%
  ungroup() %>%
  mutate(sample = factor(sample, levels = sample_levels)) %>%
  ggplot(aes(
    x = signature,
    y = sample,
    fill = exposure
  )) +
  scale_fill_viridis() +
  geom_tile()

nmf_mutation_burden_barplot <- chosen_nmf_exposures %>%
  group_by(sample) %>%
  summarise(mutation_burden = sum(exposure)) %>%
  ungroup() %>%
  mutate(sample = factor(sample, levels = sample_levels)) %>%
  ggplot(aes(
    x = sample,
    y = mutation_burden
  )) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank()
  )

plot_grid(
  nmf_exposure_heatmap + theme(legend.position = 'none'),
  nmf_mutation_burden_barplot,
  ncol = 2,
  align = 'h',
  rel_widths = c(4,2)
) %>%
  plot_grid(
    get_legend(nmf_exposure_heatmap),
    ncol = 2,
    rel_widths = c(4,1)
  )
```

# Session Info

```{r session_info}
sessionInfo()
```
